---
title: "NOAA Storm Database Exploratory Analysis"
author: "Muneeb Shahid"
date: "9/2/2020"
output:
  pdf_document: default
  html_document: default
---

## Synopsis
This document contains exploratory data analysis of NOAA Storm Database.In this analysis I have tried to find most destructive events 
in terms of population health and economic consequences. For population health effects, I took number of fatalaties and injuries caused by each event into account. For economic consequences, I took property damage and crop damage expenses into account.To compare destruction caused by harmful events, 
a bar plot was created for all of such events.

## Data Processing

Loading the Storm Data from compressed comma separated file compressed in bz2 format.

```{r, echo=TRUE, cache=TRUE}
storm_data <- read.csv("./repdata_data_StormData.csv.bz2", header=T, sep=",")
head(storm_data)
```

## Finding Most Harmful Events with respect to Population Health

Aggregating (sum) fatalities values w.r.t Event Type. 

```{r, echo=TRUE}
fatalities_per_event <- tapply(storm_data$FATALITIES, storm_data$EVTYPE, sum)
```

Aggregating (sum) injuries values w.r.t Event Type. 

```{r, echo=TRUE}
injuries_per_event <- tapply(storm_data$INJURIES, storm_data$EVTYPE, sum)
```

Taking sum of fatalities and injuries caused by each event. From now on this sum will be referred by health effects.

```{r, echo=TRUE}
health_effects_per_event <- fatalities_per_event + injuries_per_event
```

Excluding events with no health effects to find harmful events.

```{r, echo=TRUE}
health_effects_per_harmful_event <- health_effects_per_event[health_effects_per_event > 0]
```

Converting health effects per harmful event object to DataFrame type.

```{r, echo=TRUE}
health_effects_per_harmful_event_df <- data.frame(Event=names(health_effects_per_harmful_event), Health_Effects=health_effects_per_harmful_event,     row.names=NULL, stringsAsFactors=FALSE)

head(health_effects_per_harmful_event_df)
```

Caclulating average number of health effects caused by harmful events.

```{r, echo=TRUE}
harmful_events_avg_effects <- mean(health_effects_per_harmful_event_df$Health_Effects)
harmful_events_avg_effects
```

### Results

Finding most harmful events by extracting events from the dataframe having number of effects greater than average.

```{r, echo=TRUE}
hlth_efcts_per_most_hrmfl_event_df <- health_effects_per_harmful_event_df[health_effects_per_harmful_event_df$Health_Effects > harmful_events_avg_effects, ]
hlth_efcts_per_most_hrmfl_event_df
```


Plotting Health Effects against most harmful events.

```{r, echo=TRUE}
library(ggplot2)

plot <-ggplot(data=hlth_efcts_per_most_hrmfl_event_df, aes(x=Event, y=Health_Effects), ) +
  geom_bar(stat="identity") + labs(title = "Net Health Effects caused by Harmful Events") + xlab("Events") + ylab("Health Effects") + coord_flip()

plot

```

It looks like from above bar plot that Tornado is the most harmful event w.r.t population health.

## Finding Events with greatest economic consequences

Checking unique entries in property damage estimates exponents.

```{r, echo=TRUE}
unique(storm_data$PROPDMGEXP)
```
Translating exponents of property damage estimates into numbers.
This translation is recorded in new attribute.

Note: Numeric entries in exponents are translated into power of 10 and '-','+','?' values in exponents are ignored

```{r, echo=TRUE}
numeric_prop_dmg_exp <- (storm_data$PROPDMGEXP)

numeric_prop_dmg_exp <- sub("K", 1000, numeric_prop_dmg_exp)
numeric_prop_dmg_exp <- sub("h", 100, numeric_prop_dmg_exp)
numeric_prop_dmg_exp <- sub("H", 100, numeric_prop_dmg_exp)
numeric_prop_dmg_exp <- sub("B", 1000000000, numeric_prop_dmg_exp)
numeric_prop_dmg_exp <- sub("b", 1000000000, numeric_prop_dmg_exp)
numeric_prop_dmg_exp <- sub("M", 1000000, numeric_prop_dmg_exp)
numeric_prop_dmg_exp <- sub("m", 1000000, numeric_prop_dmg_exp)
numeric_prop_dmg_exp <- sub("0", 1, numeric_prop_dmg_exp)
numeric_prop_dmg_exp <- sub("1", 10, numeric_prop_dmg_exp)
numeric_prop_dmg_exp <- sub("2", 100, numeric_prop_dmg_exp)
numeric_prop_dmg_exp <- sub("3", 1000, numeric_prop_dmg_exp)
numeric_prop_dmg_exp <- sub("4", 10000, numeric_prop_dmg_exp)
numeric_prop_dmg_exp <- sub("5", 100000, numeric_prop_dmg_exp)
numeric_prop_dmg_exp <- sub("6", 1000000, numeric_prop_dmg_exp)
numeric_prop_dmg_exp <- sub("7", 10000000, numeric_prop_dmg_exp)
numeric_prop_dmg_exp <- sub("8", 100000000, numeric_prop_dmg_exp)
# replacing empty values with 0
numeric_prop_dmg_exp <- sub("", 0, numeric_prop_dmg_exp)
```

Checking unique entries in crop damage estimates exponents.

```{r, echo=TRUE}
unique(storm_data$CROPDMGEXP)
```
Translating exponents of crop damage estimates into numbers.
This translation is recorded in new attribute.

Note: Numeric entries in exponents are translated into power of 10 and '-','+','?' values in exponents are ignored

```{r, echo=TRUE}
numeric_crop_dmg_exp <- (storm_data$CROPDMGEXP)

numeric_crop_dmg_exp <- sub("K", 1000, numeric_crop_dmg_exp)
numeric_crop_dmg_exp <- sub("k", 1000, numeric_crop_dmg_exp)
numeric_crop_dmg_exp <- sub("B", 1000000000, numeric_crop_dmg_exp)
numeric_crop_dmg_exp <- sub("M", 1000000, numeric_crop_dmg_exp)
numeric_crop_dmg_exp <- sub("m", 1000000, numeric_crop_dmg_exp)
numeric_crop_dmg_exp <- sub("0", 1, numeric_crop_dmg_exp)
numeric_crop_dmg_exp <- sub("2", 100, numeric_crop_dmg_exp)
# replacing empty values with 0
numeric_crop_dmg_exp <- sub("", 0, numeric_crop_dmg_exp)


```

Creating new dataframe from original data. 
New dataframe will consist of subset of attributes from original data and newly created attributes in previous steps.

```{r, echo=TRUE}
strm_dta_fr_ecnmc_consqncs <- data.frame(storm_data$EVTYPE, storm_data$PROPDMG, storm_data$PROPDMGEXP, numeric_prop_dmg_exp, storm_data$CROPDMG, storm_data$CROPDMGEXP, numeric_crop_dmg_exp, stringsAsFactors = FALSE)

names(strm_dta_fr_ecnmc_consqncs) <- c("Event", "Prop_Dmg", "Prop_Dmg_Exp", "Numeric_Prop_Dmg_Exp", "Crop_Dmg", "Crop_Dmg_Exp", "Numeric_Crop_Dmg_Exp")

head(strm_dta_fr_ecnmc_consqncs)

```

Removing entries having '-','+','?' values in exponents.

```{r, echo=TRUE}
strm_dta_fr_ecnmc_consqncs <- strm_dta_fr_ecnmc_consqncs[strm_dta_fr_ecnmc_consqncs$Prop_Dmg_Exp != "-" | strm_dta_fr_ecnmc_consqncs$Prop_Dmg_Exp != "+" | strm_dta_fr_ecnmc_consqncs$Prop_Dmg_Exp != "?" | strm_dta_fr_ecnmc_consqncs$Crop_Dmg_Exp != "-" | strm_dta_fr_ecnmc_consqncs$Crop_Dmg_Exp != "+"| strm_dta_fr_ecnmc_consqncs$Prop_Dmg_Exp != "?",]

```

Performing Data Type conversion of attributes.

```{r, echo=TRUE}
strm_dta_fr_ecnmc_consqncs$Event <- as.character(strm_dta_fr_ecnmc_consqncs$Event)
strm_dta_fr_ecnmc_consqncs$Numeric_Crop_Dmg_Exp <- as.numeric(strm_dta_fr_ecnmc_consqncs$Numeric_Crop_Dmg_Exp)
strm_dta_fr_ecnmc_consqncs$Numeric_Prop_Dmg_Exp <- as.numeric(strm_dta_fr_ecnmc_consqncs$Numeric_Prop_Dmg_Exp)
head(strm_dta_fr_ecnmc_consqncs)
```

Multiplying Damage values with corresponding numeric exponent values and storing the result in new attributes.

```{r, echo=TRUE}
prop_dmg_expense <- strm_dta_fr_ecnmc_consqncs$Prop_Dmg * strm_dta_fr_ecnmc_consqncs$Numeric_Prop_Dmg_Exp
crop_dmg_expense <- strm_dta_fr_ecnmc_consqncs$Crop_Dmg * strm_dta_fr_ecnmc_consqncs$Numeric_Crop_Dmg_Exp

strm_dta_fr_ecnmc_consqncs$prop_dmg_expense <- prop_dmg_expense
strm_dta_fr_ecnmc_consqncs$crop_dmg_expense <- crop_dmg_expense

head(strm_dta_fr_ecnmc_consqncs)
```

Adding property damage and cost damage expenses and storing result in new attribute.

```{r, echo=TRUE}

strm_dta_fr_ecnmc_consqncs$net_dmg_expense <- strm_dta_fr_ecnmc_consqncs$prop_dmg_expense + strm_dta_fr_ecnmc_consqncs$crop_dmg_expense

head(strm_dta_fr_ecnmc_consqncs)
```

Aggregating (sum) net damage expenses w.r.t Event.

```{r, echo=TRUE}
net_dmg_expense_per_event <- tapply(strm_dta_fr_ecnmc_consqncs$net_dmg_expense, strm_dta_fr_ecnmc_consqncs$Event, sum)

## removing NA values
net_dmg_expense_per_event <- net_dmg_expense_per_event[!is.na(net_dmg_expense_per_event)]
## converting to dataframe

net_dmg_expense_per_event_df <- data.frame(Event=names(net_dmg_expense_per_event), Economic_Cost=net_dmg_expense_per_event, row.names=NULL, stringsAsFactors=FALSE)

head(net_dmg_expense_per_event_df)
```

Caclulating average economic cost across all events.

```{r, echo=TRUE}
avg_economic_cost <- mean(net_dmg_expense_per_event_df$Economic_Cost)
avg_economic_cost
```

## Results

Finding most harmful events w.r.t economic consequences by extracting events from the dataframe having economic cost greater than average.

```{r, echo=TRUE}
ecnmc_cost_per_hrmfl_event_df <- net_dmg_expense_per_event_df[net_dmg_expense_per_event_df$Economic_Cost > avg_economic_cost, ]
ecnmc_cost_per_hrmfl_event_df
```

Formatting Economic Cost in Billions

```{r, echo=TRUE}
ecnmc_cost_per_hrmfl_event_df$Economic_Cost_Billion <- round(ecnmc_cost_per_hrmfl_event_df$Economic_Cost, -3)/1000000000
ecnmc_cost_per_hrmfl_event_df
```

Plotting Economic Consequences against most harmful events.

```{r, echo=TRUE}
library(ggplot2)

ecnmc_cnsqnc_plot <-ggplot(data=ecnmc_cost_per_hrmfl_event_df, aes(x=Event, y=Economic_Cost_Billion)) +
  geom_bar(stat="identity") + labs(title = "Net Economic Cost caused by Harmful Events") + xlab("Event") + ylab("Economic Cost ($) in billions") + coord_flip()

ecnmc_cnsqnc_plot

```

From above above bar plot we can deduce that HEAVY RAIN / SEVER WEATHER has greatest economic consequences.