---
title: "NOAA Storm Database Exploratory Analysis"
author: "Muneeb Shahid"
date: "9/2/2020"
output:
  pdf_document: default
  html_document: default
---

## Loading the Storm Database

Loading the Storm Data from compressed comma separated file compressed in bz2 format.

```{r, echo=TRUE}
storm_data <- read.csv("./repdata_data_StormData.csv.bz2", header=T, sep=",")
head(storm_data)
```

## Finding Most Harmful Events with respect to Population Health

Aggregating (sum) fatalities values w.r.t Event Type. 

```{r, echo=TRUE}
fatalities_per_event <- tapply(storm_data$FATALITIES, storm_data$EVTYPE, sum)
```

Aggregating (sum) injuries values w.r.t Event Type. 

```{r, echo=TRUE}
injuries_per_event <- tapply(storm_data$INJURIES, storm_data$EVTYPE, sum)
```

Taking sum of fatalities and injuries caused by each event. From now on this sum will be referred by health effects.

```{r, echo=TRUE}
health_effects_per_event <- fatalities_per_event + injuries_per_event
```

Excluding events with no health effects to find harmful events.

```{r, echo=TRUE}
health_effects_per_harmful_event <- health_effects_per_event[health_effects_per_event > 0]
```

Converting health effects per harmful event object to DataFrame type.

```{r, echo=TRUE}
health_effects_per_harmful_event_df <- data.frame(Event=names(health_effects_per_harmful_event), Health_Effects=health_effects_per_harmful_event,     row.names=NULL, stringsAsFactors=FALSE)

head(health_effects_per_harmful_event_df)
```

Caclulating average number of health effects caused by harmful events.

```{r, echo=TRUE}
harmful_events_avg_effects <- mean(health_effects_per_harmful_event_df$Health_Effects)
harmful_events_avg_effects
```

Finding most harmful events by extracting events from the dataframe having number of effects greater than average.

```{r, echo=TRUE}
hlth_efcts_per_most_hrmfl_event_df <- health_effects_per_harmful_event_df[health_effects_per_harmful_event_df$Health_Effects > harmful_events_avg_effects, ]
hlth_efcts_per_most_hrmfl_event_df
```

### Results

Plotting Health Effects against most harmful events.

```{r, echo=TRUE}
library(ggplot2)

plot <-ggplot(data=hlth_efcts_per_most_hrmfl_event_df, aes(x=Event, y=Health_Effects)) +
  geom_bar(stat="identity") + xlab("Events") + ylab("Health Effects") + coord_flip()

plot

```

It looks like from above bar plot that Tornado is the most harmful event w.r.t population health.

## Finding Events with greatest economic consequences

### Results